---
title: "openarchaeoCollaboration"
description: |
  Advanced digital practices involving the production and manipulation of computer code are gradually becoming ubiquitous aspects of archaeological research. The relative simplicity afforded by scientific scripting languages such as R and Python, which enable custom code to be written, modified, commented, shared and reused with ease, has been a boon for fostering greater digital data manipulation skills among archaeologists. Code used to support archaeological research largely comprise custom scripts or project files that pass data through functions that parse and transform inputs according to documented routines. Base functions or functions included in generic data manipulation, analysis and visualization packages (such as those that collectively form the Tidyverse, ROpenSci and [insert Python equivalent] ecosystems, or 'off the shelf' database management systems or visualization engines) typically support a broad set of common applications, however these tools are rarely designed with archaeological use cases in mind. Recognizing the need for specialized software that more effectively suits archaeological needs, archaeologists have taken it upon themselves to create tools specifically designed for archaeological use cases. These tools operate by abstracting the factors and desired outcomes that the designers perceive to be commonly held by intended users, which happen to be archaeologists. As series of generic functions that can operate with any appropriate input data, this class of software differentiates itself from custom one-off scripts used to process and analyze data pertaining to specific research projects.

  NOTE: Refer to https://github.com/benmarwick/ctv-archaeology for an overview of common archaeological use cases, and which highlights how generic and archaeology-specific functions may be selected and applied for various applications. This is specific to R, but refer also to blogs, web resources and twitter conversations in general that discuss either how to accomplish archaeology-specific tasks using generic toolsets. Basically demonstrate that the gaps that archaeologists identify are wrapped up in and perhaps even derived from conversations about their experiences writing and manipulating code.

  This paper surveys the field of archaeological software development in order to understand what kinds of general use digital tools archaeologists are making and to identify norms regarding how software is built and maintained -- focusing specifically on collaborative aspects of software development. We use open-archaeo, a curated list of open source archaeological software and other digital resources, as the basis of this survey. This dataset is augmented with data collected from software development productivity platforms such as GitHub, which are used to host and facilitate collaborative software development work. Our analysis examines trends relating to the frequency of documented work, the distribution of contributions made by collaborating participants, how software are documented, packaged and presented, and the use of tools designed to support collaborative workflows.

  This work complements other studies that investigate scholarly use of software development productivity platforms in their work that rely on focus group sessions, questionnaires and interviews (such as [IASGE](https://investigating-archiving-git.gitlab.io/)).
  
author:
  - name: Zack Batist
    url: https://zackbatist.github.io
    affiliation: University of Toronto
  - name: Joe Roe
    url: https://joeroe.io
    affiliation: University of Copenhagen
date: "`r Sys.Date()`"
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE,
	layout = "l-body-outset"
)
```

```{r dependencies, include=FALSE}
library("tidyverse")
library("lubridate")
library("openarchaeoCollaboration")
```

## Materials and methods

We used *open-archaeo* <https://open-archaeo.info/>, a list of open source archaeological software and other digital resources.
The criteria for inclusion in open-archaeo are [...].
The search strategy for compiling it was [...].

```{r data-oarch}
read_csv("https://raw.githubusercontent.com/zackbatist/open-archaeo/master/open-archaeo.csv") %>%
  mutate(category = na_if(category, "?")) %>%
  mutate(gh_repo = str_remove(github, "https://github.com/")) ->
  oarch

oarch <- mutate(oarch, rough_category = recode(category,
                                 "R" = "Packages (R, Python, etc.)",
                                 "Web apps" = "Standalone apps",
                                 "Datasets" = "Analyses & datasets",
                                 "Tutorials" = "Lists & tutorials",
                                 "Python" = "Packages (R, Python, etc.)",
                                 "Desktop apps" = "Standalone apps",
                                 "QGIS" = "Plugins",
                                 "Lists" = "Lists & tutorials",
                                 "Mobile apps" = "Standalone apps",
                                 "Analyses" = "Analyses & datasets",
                                 "Spreadsheets" = "Forms & spreadsheets",
                                 "Command line tools" = "Standalone apps",
                                 "Plugins" = "Plugins",
                                 "ArcGis" = "Plugins",
                                 "Filemaker" = "Plugins",
                                 "ODK" = "Forms & spreadsheets",
                                 "Lisp" = "Packages (R, Python, etc.)",
                                 .default = "Other",
                                 .missing = "Other"
                                 ))
```

We augmented this with data from the GitHub API for records with an associated GitHub repository.

```{r data-github}
# Warning: slow
# TODO: map
oarch <- mutate(oarch,
                gh_langs = map(oarch$gh_repo, gh_lang),
                gh_contribs = map(oarch$gh_repo, gh_contrib),
                gh_issues = map(oarch$gh_repo, gh_issue),
                gh_comments = map(oarch$gh_repo, gh_comment),
                gh_commits = map(oarch$gh_repo, gh_commit),
)
```

## Results

### Tags and categories

```{r plot-categories}
oarch %>%
  group_by(category) %>%
  drop_na(category) %>%
  summarise(n = n()) %>%
  ggplot(aes(fct_reorder(category, n), n)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Open archaeology categories",
    caption = "Data: <https://open-archaeo.info>",
    x = NULL, y = "Number of projects"
  )
```

```{r plot-tags}
oarch %>%
  pivot_longer(c(tag1, tag2, tag3, tag4, tag5), values_to = "tag",
               values_drop_na = TRUE) %>%
  group_by(tag) %>%
  drop_na(tag) %>%
  summarise(n = n()) %>%
  ggplot(aes(fct_reorder(tag, n), n)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Open archaeology tags",
    caption = "Data: <https://open-archaeo.info>",
    x = NULL, y = "Number of projects"
  )
```

### Use of GitHub

```{r plot-gh_commits-cum}
oarch %>%
  unnest(gh_commits) %>%
  drop_na(sha) %>%
  mutate(date = floor_date(datetime, "week")) %>%
  mutate(rough_category = recode(
    rough_category,
    "Standalone apps" = "Standalone apps",
    "Packages (R, Python, etc.)" = "Packages (R, Python, etc.)",
    .default = "Other")) %>%
  group_by(rough_category, date) %>%
  summarise(n_commits = n(), .groups = "drop_last") %>%
  mutate(cat_total = sum(n_commits)) %>%
  mutate(cum_commits = cumsum(n_commits)) %>%
  ggplot(aes(date, cum_commits,
             fill = fct_reorder(rough_category, cat_total))) +
  geom_area() +
  scale_x_datetime(limits = c(ymd("2008-01-01", tz = "UTC"), NA),
                   date_breaks = "2 years",
                   date_labels = "%Y") +
  scale_fill_discrete(guide = guide_legend(reverse = TRUE)) +
  labs(
    title = "Open source archaeology",
    subtitle = "Cumulative commits in GitHub repositories",
    caption = "Data: <https://open-archaeo.info> and the GitHub API",
    x = NULL, y = "Commits", fill = NULL
  ) +
  theme(legend.position = "bottom")
```

```{r plot-gh_langs, fig.show="hold", out.width="50%"}
oarch %>%
  unnest(gh_langs) %>%
  drop_na(lang, bytes) %>%
  group_by(lang) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(-n) %>%
  slice(1:10) %>%
  ggplot(aes(fct_reorder(lang, n, .desc = TRUE), n)) +
  geom_col() +
  labs(title = "Languages used in GitHub repositories",
       subtitle = "Number of repositories using top 10 languages",
       caption = "Data: <https://open-archaeo.info> and the GitHub API",
       x = NULL, y = "Number of repositories")

oarch %>%
  unnest(gh_langs) %>%
  drop_na(lang, bytes) %>%
  group_by(lang) %>%
  summarise(bytes = sum(bytes), .groups = "drop") %>%
  arrange(-bytes) %>%
  slice(1:10) %>%
  ggplot(aes(fct_reorder(lang, bytes, .desc = TRUE), bytes)) +
  geom_col() +
  labs(title = "Languages used in GitHub repositories",
       subtitle = "Total bytes of code in top 10 languages",
       caption = "Data: <https://open-archaeo.info> and the GitHub API",
       x = NULL, y = "Bytes")
```

### Repository contributors

```{r plot-gh_contributors-histogram}
oarch %>%
  drop_na(gh_repo) %>%
  unnest(gh_contribs) %>%
  group_by(gh_repo) %>%
  summarise(n = n(), .groups = "drop") %>%
  ggplot(aes(x = n)) +
  geom_histogram(binwidth = 1) +
  labs(
    title = "Repository contributors",
    caption = "Data: <https://open-archaeo.info> and the GitHub API",
    x = "Number of contributors", y = "Number of repositories"
  )
```
